---
title: "SVM_model"
author: "Ian Stewart"
date: "2025-05-01"
output: pdf_document
---

Steps of building model:

- train the model and tune

- test model based on tuned parameters

- calculate error and accuracy

```{r, include=FALSE}
library(e1071)
library(tidyverse)
```

```{r}
# Set up multi-threading
library(parallel)

# Detect number of cores and use n-1 (leave one for system processes)
n_cores <- detectCores() - 1
cat("Utilizing", n_cores, "CPU cores for SVM training\n")

# Set OpenMP thread count for e1071/libsvm
Sys.setenv(OMP_NUM_THREADS = n_cores)
```

#Load in your own paths
```{r}
train <- read.csv("./data/train_fires10.csv")

test <- read.csv("./data/test_fires10.csv")
```


#################################################################################################################

#If already Imputed don't run!!
```{r}
train <- train %>%
  mutate(across(where(is.character), as.factor),
         owner_code = as.factor(owner_code),
         discovery_hour = as.factor(discovery_hour)) %>%
         select(-fire_size_class)
        #  burn_time_seconds = log(1+burn_time_seconds),
        #  fire_size = log(fire_size))


test <- test %>%
  mutate(across(where(is.character), as.factor),
         owner_code = as.factor(owner_code),
         discovery_hour = as.factor(discovery_hour)) %>%
         select(-fire_size_class)
        #  burn_time_seconds = log(1+burn_time_seconds),
        #  fire_size = log(fire_size))


glimpse(train)
glimpse(test)
```
####################################################################################################################

#For this code under use the variables that have been transformed and are deemed important, NOTE: get rid of the train[1:1000, ] in the models so you can train and test on the wholeeeeee data

```{r}
set.seed(123)
```

#Used cost = 1 and gamma = 0.5 based on the step above but use whatever it outputs for the whole data

```{r}
# Install and load kernlab
library(kernlab)

# Set up multi-threading (kernlab uses different settings)
options(mc.cores = detectCores() - 1)

# Tune using kernlab with better parallelization
set.seed(123)

system.time({
  # Create parameter grid
  cost_values <- c(0.1, 1, 10, 100, 1000)
  gamma_values <- c(0.5, 1, 2, 3, 4)
  param_grid <- expand.grid(cost = cost_values, sigma = gamma_values)
  
  # Store results
  accuracies <- numeric(nrow(param_grid))
  
  # Parallel tuning using kernlab
  library(doParallel)
  cl <- makeCluster(detectCores() - 1)
  registerDoParallel(cl)
  
  accuracies <- foreach(i = 1:nrow(param_grid), .combine = c, .packages = c("kernlab")) %dopar% {
    model <- ksvm(as.factor(stat_cause_descr) ~ ., 
                data = train, 
                kernel = "rbfdot",
                C = param_grid$cost[i],
                kpar = list(sigma = param_grid$sigma[i]),
                prob.model = FALSE)
    
    # Use cross-validation accuracy
    cross(model)
  }
  
  stopCluster(cl)
  
  # Find best parameters
  best_idx <- which.max(accuracies)
  best_cost <- param_grid$cost[best_idx]
  best_sigma <- param_grid$sigma[best_idx]
  
  cat("Best parameters: cost =", best_cost, ", sigma =", best_sigma, 
      ", accuracy =", accuracies[best_idx], "\n")
})
```

Best parameters: cost = 0.1 , sigma = 0.5 , accuracy = -1 
    user   system  elapsed 
   0.203    0.358 8763.288 


```{r}
svm_mod_best <- svm(stat_cause_descr ~ ., data = train, kernel = "radial",
cost = 1, gamma = 0.5)
pred <- predict(svm_mod_best, newdata = test)
table(pred, test$stat_cause_descr)

mean(pred == test$stat_cause_descr)
```
> table(pred, test$stat_cause_descr)
                   
pred                Arson Campfire Children Debris Burning Equipment Use
  Arson              2080       86      114            766           176
  Campfire             49      504       22             58            52
  Children             46        2      128             64            33
  Debris Burning     1042      250      310           3547           391
  Equipment Use        73       14       18             59           199
  Fireworks            23       14       18             23             7
  Lightning           234      517       69            245           237
  Miscellaneous       570      200      157            420           346
  Missing/Undefined    25        4       11             36            49
  Powerline             0        0        1              1             4
  Railroad              2        0        0              1             0
  Smoking               2        0        2              5             1
  Structure             0        0        0              0             0
                   
pred                Fireworks Lightning Miscellaneous Missing/Undefined
  Arson                    50       113           469                16
  Campfire                  3       111           173                 2
  Children                  7        12            48                 1
  Debris Burning           40       164           709                28
  Equipment Use            10        30           112                10
  Fireworks               102         7            25                 0
  Lightning                57      5633           673               121
  Miscellaneous            48       312          2209               136
  Missing/Undefined         1        18            91               729
  Powerline                 0         3             1                 0
  Railroad                  0         0             1                 0
  Smoking                   0         1             3                 2
  Structure                 0         0             0                 0
                   
pred                Powerline Railroad Smoking Structure
  Arson                    23       23      61        10
  Campfire                  4        3      56         0
  Children                  1        0       8        13
  Debris Burning           61      112     198        30
  Equipment Use            18        4      21         2
  Fireworks                 1        1       1         1
  Lightning                33       49     116        10
  Miscellaneous            49       24     201        10
  Missing/Undefined         3        2      13         0
  Powerline                 2        0       0         0
  Railroad                  0        6       0         0
  Smoking                   0        1       2         0
  Structure                 0        0       0         0

> mean(pred == test$stat_cause_descr)
[1] 0.565786
